---
title: "Ch2"
author: "Jill Grey Ferguson"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

```{r}
library(tidyverse)
library(sf)
library(tigris)
library(mapview)
library(leaflet)
library(censusapi)
Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab") 
knitr::opts_chunk$set(echo = F, warning = F, message = F)
smc_pop_2020 <-
  getCensus(
    name = "dec/pl",
    vintage = 2020,
    region = "block:*", 
    regionin = "state:06+county:081",
    vars = "P1_001N"
  ) %>% 
  transmute(
    block =
      paste0(state,county,tract,block),
    pop = P1_001N
  )
```

```{r}
setwd("~/Documents/GitHub/jillgrey.github.io/Assignment_1")
list.files(
)
#Now I'm going to set the computer folder where I saved all the csv's as my working directory and then list the files to make sure it sees them all. 
```

```{r}
#The chunk below is looking through all the 2017 through 2020 Quarterly PG&E electric csv files by year and quarter and then it is stacking each one and binding them all together into the variable pge_elec4years in my environment. At first, I tried setting years to 2017:2021 but I got an error:"Error: 'PGE_2021_Q3_ElectricUsageByZip.csv' does not exist in current working directory ('/Users/JillGreyFerguson/Documents/GitHub/jillgrey.github.io/Assignment 1'), since the latest available month is Q2, not Q4, so I had to do the 2021 separately and then bind them together.

years <- 2017:2020
quarters <- 1:4
type <- "Electric"

pge_elec4years <- NULL

for(year in years) {
  for(quarter in quarters) {
    
    filename <- 
      paste0(
        "PGE_",
        year,
        "_Q",
        quarter,
        "_",
        type,
        "UsageByZip.csv"
      )
  
    print(filename)
    
    temp <- read_csv(filename)
    
    pge_elec4years <- rbind(pge_elec4years,temp)
  }
}
years <- 2021
quarters <- 1:2
type <- "Electric"

pge_elec_this_year <- NULL

for(year in years) {
  for(quarter in quarters) {
    
    filename <- 
      paste0(
        "PGE_",
        year,
        "_Q",
        quarter,
        "_",
        type,
        "UsageByZip.csv"
      )
  
    print(filename)
    
    temp <- read_csv(filename)
    
    pge_elec_this_year <- rbind(pge_elec_this_year,temp)
  
  }
}
pge_electric <- rbind(pge_elec4years,pge_elec_this_year)
saveRDS(pge_electric, "pge_electric.rds")
```
```

```{r}
bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME %in% bay_county_names)
```

```{r}
ggplot(bay_counties) + geom_sf()
```

```{r}
bay_cbgs_clip <- st_read("https://opendata.arcgis.com/datasets/037fc1597b5a4c6994b89c46a8fb4f06_0.geojson")
```

```{r}
ca_cities <- places("CA", cb = T, progress_bar = FALSE)
```

```{r}
bay_cities <- ca_cities[bay_counties, ]

mapview(bay_counties, alpha.regions = 0) + mapview(bay_cities)
```

```{r}
bay_cities_within <-
  ca_cities %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_set_geometry(NULL) %>% 
  left_join(ca_cities %>% select(GEOID)) %>% 
  st_as_sf()

mapview(bay_counties, alpha.regions = 0) + mapview(bay_cities_within, label = "NAME")
```
#```{r}
#bay_cities_within <-
#  ca_cities[which(ca_cities$GEOID %in% st_centroid(ca_cities)[bay_counties, ]$GEOID), ]
#```

```{r}
leaflet() %>% 
  addTiles() %>% 
  addPolygons(
    data = bay_counties,
    fill = F,
    weight = 2,
    label = ~NAME
  ) %>%
  addPolygons(
    data = bay_cities %>% 
      filter(!GEOID %in% bay_cities_within$GEOID),
    color = "red",
    label = ~NAME
  ) %>% 
  addPolygons(
    data = bay_cities_within,
    color = "green",
    label = ~NAME
  )
```

```{r}
bay_cbgs <- block_groups("CA", bay_county_names[1:9], cb = T, progress_bar = F)
```

```{r}
bay_cbgs <- 
  bay_county_names %>% 
  map_dfr(function(county) {
    block_groups("CA", county, cb = T, progress_bar = F)
  })
```

```{r}
ggplot(bay_cbgs_clip)+geom_sf()
```

```{r}
usa_zips <- 
  zctas(cb = T, progress_bar = F)

bay_zips <-
  usa_zips %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_set_geometry(NULL) %>% 
  left_join(usa_zips %>% select(GEOID10)) %>% 
  st_as_sf()

pge_residential_electric <-
  pge_electric %>% 
  filter(CUSTOMERCLASS == "Elec- Residential") %>% 
  mutate(
    ZIPCODE = ZIPCODE %>% as.character()
  ) %>% 
  group_by(ZIPCODE) %>% 
  summarize(
    TOTALKWH = sum(TOTALKWH, na.rm = T)
  ) %>% 
  right_join(
    bay_zips %>% select(GEOID10),
    by = c("ZIPCODE" = "GEOID10")
  ) %>% 
  st_as_sf() %>% 
  st_transform(4326)
```

```{r}
res_pal <- colorNumeric(
  palette = "Reds",
  domain = 
    pge_residential_electric$TOTALKWH
)

leaflet() %>% 
  addTiles() %>% 
  addPolygons(
    data = pge_residential_electric,
    fillColor = ~res_pal(TOTALKWH),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.5,
    weight = 1,
    label = ~paste0(
      round(TOTALKWH), 
      " kWh total in ",
      ZIPCODE
    ),
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  ) %>% 
  addLegend(
    data = pge_residential_electric,
    pal = res_pal,
    values = ~TOTALKWH,
    title = "Total Residential<br>kWh, 2020"
  )
```

