---
title: "Assignment 1: PG&E"
author: "Jill Grey Ferguson"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---


```{r}
library(tidyverse)
knitr::opts_chunk$set(echo = F, warning = F, message = F)
#The first chunk is to make sure I have all the packages I need for this assignment
```

```{r}
setwd("~/Documents/GitHub/jillgrey.github.io/Assignment_1")
list.files(
)
#Now I'm going to set the computer folder where I saved all the csv's as my working directory and then list the files to make sure it sees them all. 
```

```{r}
#The chunk below is looking through all the 2017 through 2020 Quarterly PG&E electric csv files by year and quarter and then it is stacking each one and binding them all together into the variable pge_elec4years in my environment. At first, I tried setting years to 2017:2021 but I got an error:"Error: 'PGE_2021_Q3_ElectricUsageByZip.csv' does not exist in current working directory ('/Users/JillGreyFerguson/Documents/GitHub/jillgrey.github.io/Assignment 1'), since the latest available month is Q2, not Q4, so I had to do the 2021 separately and then bind them together.

years <- 2017:2020
quarters <- 1:4
type <- "Electric"

pge_elec4years <- NULL

for(year in years) {
  for(quarter in quarters) {
    
    filename <- 
      paste0(
        "PGE_",
        year,
        "_Q",
        quarter,
        "_",
        type,
        "UsageByZip.csv"
      )
  
    print(filename)
    
    temp <- read_csv(filename)
    
    pge_elec4years <- rbind(pge_elec4years,temp)
  }
}
years <- 2021
quarters <- 1:2
type <- "Electric"

pge_elec_this_year <- NULL

for(year in years) {
  for(quarter in quarters) {
    
    filename <- 
      paste0(
        "PGE_",
        year,
        "_Q",
        quarter,
        "_",
        type,
        "UsageByZip.csv"
      )
  
    print(filename)
    
    temp <- read_csv(filename)
    
    pge_elec_this_year <- rbind(pge_elec_this_year,temp)
  
  }
}
pge_electric <- rbind(pge_elec4years,pge_elec_this_year)
saveRDS(pge_electric, "pge_electric.rds")
```

```{r}
#I'm now checking my work to see if all years, including 2021, do in fact show up here and they do, so I commented it out so it doesn't take up my whole html file 
# ```{r}
# View(pge_electric)
# ```
#The next chunk is doing the same for gas

years <- 2017:2020
quarters <- 1:4
type <- "Gas"

pge_gas4years <- NULL

for(year in years) {
  for(quarter in quarters) {
    
    filename <- 
      paste0(
        "PGE_",
        year,
        "_Q",
        quarter,
        "_",
        type,
        "UsageByZip.csv"
      )
  
    print(filename)
    
    temp <- read_csv(filename)
    
    pge_gas4years <- rbind(pge_gas4years,temp)
  
  }
}
years <- 2021
quarters <- 1:2
type <- "Gas"

pge_gas_this_year <- NULL

for(year in years) {
  for(quarter in quarters) {
    
    filename <- 
      paste0(
        "PGE_",
        year,
        "_Q",
        quarter,
        "_",
        type,
        "UsageByZip.csv"
      )
  
    print(filename)
    
    temp <- read_csv(filename)
    
    pge_gas_this_year <- rbind(pge_gas_this_year,temp)
  
  }
}
pge_gas <- rbind(pge_gas4years,pge_gas_this_year)
saveRDS(pge_gas, "pge_gas.rds")
```


```{r}
#Also checking the gas data
# ```{r}
# View(pge_gas)
# ```
#The next chunk filters to get the customer classes we care about for electric and then I realized I can click in the environment to view the data rather than having to use the view function each time.  


pge_electric_filter <- filter(pge_electric, CUSTOMERCLASS %in% c("Elec- Residential","Elec- Commercial"))
```


```{r}
#The next chunk does the same but for gas and I checked in the environment.

pge_gas_filter <- filter(pge_gas, CUSTOMERCLASS %in% c("Gas- Residential","Gas- Commercial"))
```


```{r}
#Now I'm thinking that I need to put gas and electric together but the columns aren't all the same (the gas and electric units are different) so I think I need to make them both the same units before I bind. After converting units and changing the name, I  viewed the environment to confirm that it did create a column with the converted units. 

pge_electric_filter_convert_units <- mutate (pge_electric_filter, TOTALKBTUs=TOTALKWH*3412.14/1000)
```

```{r}
#And now the same for gas

pge_gas_filter_convert_units <- mutate (pge_gas_filter, TOTALKBTUs=TOTALTHM*99976.1/1000)
```

```{r}
#Before I bind, I want to select only the columns I care about for electric. 

  pge_electric_final <- pge_electric_filter_convert_units %>% select(
    -c(ZIPCODE, COMBINED, TOTALCUSTOMERS,TOTALKWH,AVERAGEKWH))
```



```{r}
#Now I will do the same for gas. 

  pge_gas_final <- pge_gas_filter_convert_units %>% select(
    -c(ZIPCODE, COMBINED, TOTALCUSTOMERS,TOTALTHM,AVERAGETHM))
```

```{r}
#Now I can finally bind the gas and electric.

pge_gas_electric <- rbind(pge_electric_final,pge_gas_final)
saveRDS(pge_gas_electric, "pge_gas_electric.rds")
```

```{r}
#Now I'm going to attempt to make a chart for residential, which requires removing the commercial rows for now and also changing the date format so that it splits into 54 months.

pge_final_residential <-
  pge_gas_electric %>% 
  filter(
    CUSTOMERCLASS %in% 
      c(
        "Elec- Residential", 
        "Gas- Residential"
      )
  ) %>% 
  group_by(
    MONTH, 
    CUSTOMERCLASS, 
    YEAR
  ) %>% 
  summarize(
    TOTALKBTUs = sum(TOTALKBTUs, na.rm = T)
  ) %>% 
  mutate(
    DATE = 
      paste(
        YEAR,
        MONTH, 
        "01",
        sep="-"
      ) %>% as.Date()
  )
head(pge_final_residential)
```

```{r}
#Now I'm attempting the residential plot. 

pge_final_residential_chart <- pge_final_residential %>% 
  filter(CUSTOMERCLASS %in%   
      c(
        "Elec- Residential", 
        "Gas- Residential"
      )) %>% 
  ggplot() +
  geom_line(
    aes(
      x = DATE %>% as.Date(),
      y = TOTALKBTUs,
      color = CUSTOMERCLASS,
    )) + 
  labs(
    x = "Year",
    y = "Total kBTUs",
    title = "PG&E Territory Residential Gas and Electricity Usage, 2017 - 2021",
    color = "Fuel Type",
   # theme(color.position="bottom")
  ) +
  theme(
    legend.position = "bottom"
  )
pge_final_residential_chart
```


```{r}
#Now I'm going to repeat but for commercial 

pge_final_commercial <-
  pge_gas_electric %>% 
  filter(
    CUSTOMERCLASS %in% 
      c(
        "Elec- Commercial", 
        "Gas- Commercial"
      )
  ) %>% 
  group_by(
    MONTH, 
    CUSTOMERCLASS, 
    YEAR
  ) %>% 
  summarize(
    TOTALKBTUs = sum(TOTALKBTUs, na.rm = T)
  ) %>% 
  mutate(
    DATE = 
      paste(
        YEAR,
        MONTH, 
        "01",
        sep="-"
      ) %>% as.Date()
  )
```


```{r}
#On to the commercial plot 

pge_final_commercial_chart <- pge_final_commercial %>% 
  filter(CUSTOMERCLASS %in%   
      c(
        "Elec- Commercial", 
        "Gas- Commercial"
      )) %>% 
  ggplot() +
  geom_line(
    aes(
      x = DATE %>% as.Date(),
      y = TOTALKBTUs,
      color = CUSTOMERCLASS,
    ))+ 
  labs(
    x = "Year",
    y = "Total kBTUs",
    title = "PG&E Territory Commercial Gas and Electricity Usage, 2017 - 2021",
    color = "Fuel Type",
  )  +
  theme(
    legend.position = "bottom")
pge_final_commercial_chart
```



```{r}
#Now, to zoom in on the COVID-19 pandemic, I need to repeat both graphs but for 2019-2020 starting with Residential 

pge_COVID_residential <-
  pge_gas_electric %>%
  filter(
    CUSTOMERCLASS %in%
      c(
        "Elec- Residential",
        "Gas- Residential"
      )
    )%>% 
  filter(
    YEAR %in%   
      c(2019,
         2020
      )
  ) %>% 
  group_by(
    MONTH, 
    CUSTOMERCLASS, 
    YEAR
  ) %>% 
  summarize(
    TOTALKBTUs = sum(TOTALKBTUs, na.rm = T)
  )%>%
  mutate(
    DATE =
      paste(
        YEAR,
        MONTH,
        "01",
        sep="-"
      ) %>% as.Date()
)
```

```{r}
pge_COVID_residential_chart <- pge_COVID_residential %>% 
  filter(CUSTOMERCLASS %in%   
      c(
        "Elec- Residential", 
        "Gas- Residential"
      )
      ) %>% 
  ggplot() +
  scale_x_discrete(
    limits = c(
      "Jan",
      "Feb",
      "Mar",
      "Apr",
      "May",
      "Jun",
      "Jul",
      "Aug",
      "Sep",
      "Oct",
      "Nov",
      "Dec"
    )
  ) + 
  scale_linetype_manual(values = c('2019' = 2,'2020' = 1))+
   geom_line(
    aes(
      x = MONTH ,
      y = TOTALKBTUs,
      color = CUSTOMERCLASS,
      linetype = YEAR %>% factor()
    )
  ) + 
  labs(
    x = "MONTH",
    y = "Total kBTUs",
    title = "PG&E Territory Residential Gas and Electricity Usage, COVID-19 2019 vs 2020",
    color = "Fuel Type",
    linetype = "YEAR"
  ) +
  theme(
    legend.position = "bottom"
  )
pge_COVID_residential_chart
```

### Interpretation of COVID-19's impact on PG&E's residential energy usage (2019 vs 2020):

### Assumptions and Caveats 
* 2019 is pre-COIVD period
* 2020 is post-COVID period 
* We would need to conduct weather normalization on pre-COVID home energy usage (2019) and post-COVID home energy usage to really isolate COVID-19's impact. 
* Energy demand graphs cannot differentiate between reduced demand due to weather vs curtailment of energy usage due to lack of economic resources, the latter of which could be considered COVID related since COVID was an exacerbating factor for households. We'd need to investigate PG&E's disconnection rate, arrearages, and late payments to dive into this aspect. 

### Observations 
* There doesn't appear to be a significant difference in home energy usage pre- and post- COVID with respect to both gas and electric usage.
* In almost every month, post-COVID gas and electric usage/demand is higher than before COVID, suggesting that the year of lockdown did lead to some increased residential usage, barring differences in climate. 
* The only exception is gas and electric usage in the month of February which was higher before COVID than after, but again this could be because weather in 2020 was more mild than in 2019. If this were the case, people would be cranking their heat pumps or burning more gas for heat in 2019 which created more overall demand than did the lockdown during a warm February. 
* With respect to home electricity usage, the general shape of the demand curve is consistent pre- and post- COVID with peaks in August and December, presumably due to an increase in electric air conditioners and non-fuel based heating, respectively. 
* Gas usage pre- and post- COVID also follow the same general shape, peaking in winter as expected. The other  difference beyond February is 2020 gas usage in April, which is higher than 2019 by more kBTUs than in any other month beyond February. But again, April 2020 could have been cooler than in 2019.


```{r}
#Now, to plot Commercial 

pge_COVID_commercial <-
  pge_gas_electric %>%
  filter(
    CUSTOMERCLASS %in%
      c(
        "Elec- Commercial",
        "Gas- Commercial"
      )
    )%>% 
  filter(
    YEAR %in%   
      c(2019,
         2020
      )
  ) %>% 
  group_by(
    MONTH, 
    CUSTOMERCLASS, 
    YEAR
  ) %>% 
  summarize(
    TOTALKBTUs = sum(TOTALKBTUs, na.rm = T)
  )%>%
  mutate(
    DATE =
      paste(
        YEAR,
        MONTH,
        "01",
        sep="-"
      ) %>% as.Date()
)
```

```{r}
pge_COVID_commercial_chart <- pge_COVID_commercial %>% 
  filter(CUSTOMERCLASS %in%   
      c(
        "Elec- Commercial", 
        "Gas- Commercial"
      )
      ) %>% 
  ggplot() +
  scale_x_discrete(
    limits = c(
      "Jan",
      "Feb",
      "Mar",
      "Apr",
      "May",
      "Jun",
      "Jul",
      "Aug",
      "Sep",
      "Oct",
      "Nov",
      "Dec"
    )
  ) + 
  scale_linetype_manual(values = c('2019' = 2,'2020' = 1))+
   geom_line(
    aes(
      x = MONTH ,
      y = TOTALKBTUs,
      color = CUSTOMERCLASS,
      linetype = YEAR %>% factor()
    )
  ) + 
  labs(
    x = "MONTH",
    y = "Total kBTUs",
    title = "PG&E Territory Commercial Gas and Electricity Usage, COVID-19 2019 vs 2020",
    color = "Fuel Type",
    linetype = "YEAR"
  ) +
  theme(
    legend.position = "bottom"
  )
pge_COVID_commercial_chart
```

### Interpretation of COVID-19's impact on PG&E's commercial energy usage (2019 vs 2020):

### Assumptions and Caveats 
* 2019 is pre-COIVD period
* 2020 is post-COVID period 
* We would need to conduct weather normalization on pre-COVID commercial energy usage (2019) and post-COVID commercial energy usage to really isolate COVID-19's impact. 
* Data source is PG&E in California, which, was more precautions in terms of timing and extent of lockdown policies than say Southern U.S. states. The next graph could compare to GA Power energy usage, for example, although I didn't have time to follow this through.  

### Observations 
* There is a much larger observable difference in commercial energy usage pre- and post- COVID than in the residential sector. 
* In just about every month, commercial energy usage for both gas and electric were higher pre-COVID than post-COVID, as expected. 
* Barring weather variations, this suggests that lockdowns and work from home orders caused a significant drop in energy usage in restaurants, retail stores, grocery stores, hotels, and more as these businesses either shut down or had their facility managers significantly reduce operation. This is expected since so many businesses closed completely or operated at reduced hours and shifted to take out rather than dine in.   
* If repeated to incldue 2021, we should see an uptick in commerical energy usage in PG&E territory since businesses (and universities) have started to re-open. 


```{r}
#Let's try that below

pge_COVID_commercial_2021 <-
  pge_gas_electric %>%
  filter(
    CUSTOMERCLASS %in%
      c(
        "Elec- Commercial",
        "Gas- Commercial"
      )
    )%>% 
  filter(
    YEAR %in%   
      c(2019,
         2020,
        2021
      )
  ) %>% 
  group_by(
    MONTH, 
    CUSTOMERCLASS, 
    YEAR
  ) %>% 
  summarize(
    TOTALKBTUs = sum(TOTALKBTUs, na.rm = T)
  )%>%
  mutate(
    DATE =
      paste(
        YEAR,
        MONTH,
        "01",
        sep="-"
      ) %>% as.Date()
)
```

```{r}
pge_COVID_commercial_2021_chart <- pge_COVID_commercial_2021 %>% 
  filter(CUSTOMERCLASS %in%   
      c(
        "Elec- Commercial", 
        "Gas- Commercial"
      )
      ) %>% 
  ggplot() +
  scale_x_discrete(
    limits = c(
      "Jan",
      "Feb",
      "Mar",
      "Apr",
      "May",
      "Jun",
      "Jul",
      "Aug",
      "Sep",
      "Oct",
      "Nov",
      "Dec"
    )
  ) + 
  scale_linetype_manual(values = c('2019' = 2,'2020' = 1, '2021'=3))+
   geom_line(
    aes(
      x = MONTH ,
      y = TOTALKBTUs,
      color = CUSTOMERCLASS,
      linetype = YEAR %>% factor()
    )
  ) + 
  labs(
    x = "MONTH",
    y = "Total kBTUs",
    title = "PG&E Territory Commercial Gas and Electricity Usage, COVID-19 2019-2021",
    color = "Fuel Type",
    linetype = "YEAR"
  ) +
  theme(
    legend.position = "bottom"
  )
pge_COVID_commercial_2021_chart
```

### Interpretation of COVID-19's impact on PG&E's commercial energy usage (2019 through 2021):

### Assumptions
* 2021 represents the after COVID period 
* 2019 is still pre-COVID 
* 2020 is now the "during COVID" period since 2021 is now included 

### Observations
* As predicted, both electric and gas consumption after COVID (2021) have increased compared to the during COVID (2020) lockdowns. 
* However, after COVID electricity usage is also significantly higher than pre-COVID (2019) levels, indicating that business operation did not simply return to "normal" but rather ramped-up to make up for "lost time."
* After COVID gas usages also increases above pre-COVID usage in March and part of April, but pre-COVID gas usage is higher in the other months. This could indicate a cooler 2021 spring than 2019 and/or more business activity that required more fuel.  
